---
title: "Methylation"
format: html
editor: visual
knitr:
  opts_chunk: 
    cache : true
    cache.comments: false
    cache.lazy: false
    collapse: true
    warning: false
    message: false
    echo: false
    strip.white: true
    comment: "#>"
    fig.path: "../figures/"
    results: "asis"
    tidy: "styler"
    dev: "pdf"
    dev.args: list(pdf = list(colormodel = 'cmyk'), 
                  tiff = list(compression = 'lzw'))
    dpi: 600
    fig.width: 8
    fig.height: 8
    concordance: true
    global.par: true
    R.options:
      scipen: 999
      huxtable.long_minus: true
---

Side project to look for relation between MAO-A methylation data and nutrition

```{r load-data}
library(tidyverse)
library(here)
library(skimr)
library(ggsci)
library(ggdark)
library(correlation)
library(GGally)
library(lme4)
library(lmerTest)
library(buildmer)
library(geepack)
library(ggstatsplot)
library(sjPlot)
library(sjlabelled)
library(labelled)
library(gt)
library(gtsummary)
library(flextable)
library(gvlma)
library(mgcv)
library(lmtest)
library(sandwich)
library(parameters)
library(estimatr)
library(Cairo)

df_meth <- readr::read_csv(here("raw_data", "methylation.csv"), col_types = "innnff")

df_eliktu <- readr::read_csv(here("raw_data", "df_lapsed.csv")) %>%
  dplyr::filter(!rowSums(is.na(.)) >= 64) %>% 
  dplyr::rowwise() %>% 
  dplyr::mutate(aImp = sum(c(AMIS3, (6-AMIS7), AMIS11, AMIS15, AMIS19, AMIS23,
                             AMIS2, (6-AMIS6), AMIS10, AMIS14, AMIS18, AMIS22)),
                mImp = sum(c(AMIS1, AMIS5, AMIS9, AMIS13, AMIS17, (6-AMIS21),
                             AMIS4, AMIS12, AMIS16, AMIS20, AMIS24))) %>%
  dplyr::ungroup() %>%
  dplyr::select(-c(AMIS1:AMIS24)) %>% 
  dplyr::filter(!(kood == 1824 & age == 18)) %>% 
  dplyr::filter(!(kood == 1157 & age == 33)) %>% 
  dplyr::filter(!(kood == 1307 & age == 18)) %>%
  dplyr::select(-c(sugu:kohort), -c(aImp:mImp), -LDL, -Wmaxkg)
  
df_meth_tall <- df_meth |> tidyr::pivot_longer(cols = c(-kood, -MAOA.uVNTR, -Maltreatment.18),
                                               names_to = "age",
                                               values_to = "CpG3.methylation",
                                               names_prefix = "CpG3.methylation.",
                                               names_transform = as.integer)

df <- df_meth_tall |> dplyr::left_join(df_eliktu, by = c("kood", "age")) %>%
  dplyr::filter(!rowSums(is.na(.)) >= 34)
  

rowSums(is.na(df))

# saveRDS(df, here("data", "eliktu_lapsed_methylation_data.rds"), compress='xz')
# readr::write_csv2(df, here("data", "eliktu_lapsed_methylation_data.csv"))
```

```{r methyl-data}
skim(df_meth[, -c(1, 5)])

```

```{r correlations}

correlation(df_meth[, -c(1, 5)])
  
```

```{r ggpairs}
#| fig.height: 10

ggpairs(df_meth[, 2:4], mapping = aes(color = df_meth$MAOA.uVNTR))
```

```{r model-prep}
df_scaled  <- df %>% 
  dplyr::mutate(kood = factor(kood), age_cent = (age - 18), age = NULL) %>% 
  dplyr::mutate(across(CpG3.methylation:Alco, ~ c(scale(., scale = TRUE)))) %>%
  dplyr::select(-Maltreatment.18)

labels_scaled <- list(age_cent = "Age, years",
               Calcium = "Calcium, mg", Iodine = "Iodine, µg",
               Iron = "Iron, mg", Magnesium = "Magnesium, mg",
               Manganese = "Manganese, mg", Phosphorus = "Phosphorus, mg",
               Potassium = "Potassium, mg", Selenium  = "Selenium, µg",
               Sodium = "Sodium, mg", Zink = "Zinc, mg",
               kcal = "Daily energy intake, kcal/day",
               BMI = "BMI, kg/m^2",
               Carb = "Carbohydrates, g", Cerealprod = "Cereal products, g",
               CHL = "Cholesterol, mmol/l", Eggs = "Eggs, g",
               Fatsg = "Fats, g", Fish = "Fish, g", Folate = "Folate, µg",
               FruitsBerries = "Fruits & berries, g",
               HDL = "HDL-cholesterol, mmol/l",
               HOMA = "HOMA, units", MAOA.uVNTR = "MAOA uVNTR",
               Lipid = "Lipids, g", Meat = "Meat, g", Milk = "Milk, g",
               Niacin = "Niacin, mg", Protein = "Proteins, g",
               SugSweets = "Sugar & sweets, g", Veget = "Vegetables, g",
               VitA = "Vitamin A, µg", VitB1 = "Vitamin B1, mg",
               VitB12 = "Vitamin B12, µg", VitB2 = "Vitamin B2, mg",
               VitB6 = "Vitamin B6, mg", VitC = "Vitamin C, mg",
               VitD = "Vitamin D, µg", VitE = "Vitamin E, mg",
               Alco = "Pure alcohol, g",
               CpG3.methylation = "MAOA CpG3 methylation")

labels <- list(age = "Age, years",
               Calcium = "Calcium, mg", Iodine = "Iodine, µg",
               Iron = "Iron, mg", Magnesium = "Magnesium, mg",
               Manganese = "Manganese, mg", Phosphorus = "Phosphorus, mg",
               Potassium = "Potassium, mg", Selenium  = "Selenium, µg",
               Sodium = "Sodium, mg", Zink = "Zinc, mg",
               kcal = "Daily energy intake, kcal/day",
               BMI = "BMI, kg/m^2",
               Carb = "Carbohydrates, g", Cerealprod = "Cereal products, g",
               CHL = "Cholesterol, mmol/l", Eggs = "Eggs, g",
               Fatsg = "Fats, g", Fish = "Fish, g", Folate = "Folate, µg",
               FruitsBerries = "Fruits & berries, g",
               HDL = "HDL-cholesterol, mmol/l",
               HOMA = "HOMA, units", MAOA.uVNTR = "MAOA uVNTR",
               Lipid = "Lipids, g", Meat = "Meat, g", Milk = "Milk, g",
               Niacin = "Niacin, mg", Protein = "Proteins, g",
               SugSweets = "Sugar & sweets, g", Veget = "Vegetables, g",
               VitA = "Vitamin A, µg", VitB1 = "Vitamin B1, mg",
               VitB12 = "Vitamin B12, µg", VitB2 = "Vitamin B2, mg",
               VitB6 = "Vitamin B6, mg", VitC = "Vitamin C, mg",
               VitD = "Vitamin D, µg", VitE = "Vitamin E, mg",
               Alco = "Pure alcohol, g",
               CpG3.methylation = "MAOA CpG3 methylation")

labels_vec <- unlist(labels_scaled)
var_label(df_scaled) <- labels_scaled
var_label(df) <- labels

# readr::write_csv2(df_scaled, here("data", "eliktu_lapsed_methylation_data_scaled.csv"))
```

```{r formula}

fmla <- reformulate(c(names(df_scaled)[-c(1, 3)], "(1 + age_cent | kood)"), names(df_scaled)[3])


```

**Finding the maximal *feasible* model & doing stepwise elimination from it**

The selected formula after the backward elimination `CpG3.methylation ~ 1 + Zink + age_cent`

The criterion for variable elimination is the significance of the change in log-likelihood (`LRT`: terms which provide lower chi-square values are considered more important)

`lmer` models do not converge even with only few variables, therefore `lm` was used.

```{r lmer}
m1 <- buildmer(fmla, data = df_scaled, 
         buildmerControl = buildmerControl(direction = 'backward',
          args = list(control = lmerControl(optimizer = 'bobyqa'))))

```

```{r model-summary}
summary(m1)

```

Using AIC as a variable selection criterion

```{r lmer2}
fmla2 <- reformulate(c(names(df_scaled)[-c(1, 3)], "(1 | kood)"), names(df_scaled)[3])
m2 <- buildmer(fmla2, data = df_scaled, 
         buildmerControl = buildmerControl(direction = c('order', 'backward'),
                                           crit = 'AIC',
          args = list(control = lmerControl(optimizer = 'bobyqa'))))
summary(m2)
```
Same as model m2 but I am using the change in BIC score as criterion



glmmTMB

```{summary(m3){r glmmTMB}

m3 <- buildglmmTMB(fmla2, data = df_scaled, 
         buildmerControl = buildmerControl(direction = c('order', 'backward'),
                                           crit = 'AIC'))
```

There was no significant interaction between age and zinc.Then I added MAO-A genotype and maltreatment factors.

```{r age-zink}

fmla3 <- as.formula("CpG3.methylation ~ Zink + age_cent + Maltreatment.18 * MAOA.uVNTR")

m4 <- lm(fmla3, data = df_scaled)
summary(m4)
```

**Generalised estimating equations (GEE)**

GEE estimates population average effects.

```{r gee}

df_scaled2 <- df_scaled |> dplyr::mutate(age_cent = age_cent + 3L)

fmla4 <- formula(CpG3.methylation ~ age_cent + BMI + CHL + Fatsg + Protein + FruitsBerries + Alco + Calcium + Iodine + Iron + Manganese + Phosphorus + Potassium + Zink + VitB1 + VitE)

fmla4_MAOA <- formula(CpG3.methylation ~ age_cent + BMI + CHL + Fatsg + Protein + FruitsBerries + Alco + Calcium + Iodine + Iron + Manganese + Phosphorus + Potassium + Zink + VitB1 + VitE + MAOA.uVNTR)

fmla4
m5 <- geeglm(fmla4, data = na.omit(df_scaled), id = kood, family = gaussian, corstr= "ex", scale.fix = FALSE)
summary(m5)
```

robust regression

```{r lm-robust}

m2_robust <- lm_robust(fmla4, data = df_scaled, se_type = "HC2")
summary(m2_robust)

m2_robust_MAOA <- lm_robust(fmla4_MAOA, data = df_scaled, se_type = "HC2")
summary(m2_robust_MAOA)
```


MAO-A genotype and maltreatment at age 18

```{r maltr-18}

df_18 <- df |> dplyr::filter(age == 18L)

grouped_ggbetweenstats(
  data             = df_18,
  x                = Maltreatment.18,
  y                = CpG3.methylation,
  grouping.var     = MAOA.uVNTR,
  outlier.tagging  = TRUE,
  outlier.label    = kood,
  outlier.coef     = 2,
  ggsignif.args    = list(textsize = 4, tip_length = 0.01),
  p.adjust.method  = "bonferroni",
  palette          = "default_jama",
  package          = "ggsci",
  plotgrid.args    = list(nrow = 1),
  annotation.args  = list(title = "Differences in CpG3 methylation by maltreatment status at age 18 for MAO-A genotypes")
)

```
check regression model assumptions. It was not a perfect, but there too few observations to try 

```{r lm-check}

gvlma(m2@model)

# calculate robust standard errors for model coefficients
 
coeftest(m2@model, vcov = vcovHC(m2@model, type = 'HC0'))

```

```{r exp-plots}

df_tall <- df_scaled %>% pivot_longer(-c(kood, CpG3.methylation, MAOA.uVNTR),
                                      names_to = "measure",
                                      values_to = "value",
                                      values_drop_na = TRUE)

df_tall %>% ggplot(aes(x = value, y = CpG3.methylation)) +
  geom_point(alpha = 0.2, size = 0.2) +
  geom_smooth() +
  facet_wrap(~ measure, scales = "free") + scale_color_viridis_d()
```



```{r coef-plot}
theme_set(theme_gray())

model_plot <- plot_models(list(GEE = m5, "robust LM" = m2_robust),
           title = "MAOA CpG3 methylation", vline.color = "#63c132", 
           m.labels = c("GEE", "robust LM"), p.shape = TRUE,
           show.p = TRUE, p.threshold = c(0.05, 0.01, 0.001),
           legend.title = "Model", show.legend = TRUE,
           show.values = FALSE, axis.labels = labels_vec,
           spacing = 0.6, value.size = 1.8, line.size = 0.9,
           dot.size = 3, colors = c("#E41A1C", "#6BAED6")) +
  theme(text = element_text(vjust = 0.2, size = 14),
        axis.line = element_line(color="black"), plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5)) + 
  labs(y = "Regression coefficients") +
  guides(shape = guide_legend(override.aes = list(size = 3, color = "black")))

model_plot$data[["p.stars"]] <- fct_recode(model_plot$data[["p.stars"]],
           "<0.05" = "*", "<0.01" = "**", "<0.001" = "***")

model_plot

fpath <- here("figures", "methylation-models-plot-plevels-dark-red2.eps")
ggsave(fpath, model_plot, dpi = 600, width = 8, height = 4.5, device = "eps")
```

```{r model-summaries}
fpath <- here("tables", "methylation.html")
tab_model(list(GEE = m5, "robust LM" = m2_robust), show.intercept = FALSE, 
          dv.labels = c("GEE", "robust LM"), show.ngroups = FALSE, file = fpath)

```

```{r summary-table}
paper_suppl_tbl <- df %>% select(-c(kood, Maltreatment.18)) %>% 
  tbl_summary(by = "age", missing_text = "Missing observations") 
paper_suppl_tbl

path_doc <- here("tables", "summary_table_suppl.docx")
ft <- paper_suppl_tbl %>% as_flex_table()
ft <- font(ft, fontname = "Times New Roman", part = c("all"))
ft <- fontsize(ft, size = 8)
ft <- fontsize(ft, size = 10, part = c("header"))
ft <- fontsize(ft, size = 10, part = c("footer"))
ft <- line_spacing(ft, space = 1) %>% autofit() %>% fit_to_width(max_width = 11)
save_as_docx(ft, path = path_doc)

```

