---
title: "Methylation"
format: html
editor: visual
knitr:
  opts_chunk: 
    cache : true
    cache.comments: false
    cache.lazy: false
    collapse: true
    warning: false
    message: false
    echo: false
    strip.white: true
    comment: "#>"
    fig.path: "../figures/"
    results: "asis"
    tidy: "styler"
    dev: "pdf"
    dev.args: list(pdf = list(colormodel = 'cmyk'), 
                  tiff = list(compression = 'lzw'))
    dpi: 600
    fig.width: 8
    fig.height: 8
    concordance: true
    global.par: true
    R.options:
      scipen: 999
      huxtable.long_minus: true
---

Side project to look for relation between MAO-A methylation data and nutrition

```{r load-data}
library(tidyverse)
library(here)
library(skimr)
library(correlation)
library(GGally)
library(lme4)
library(lmerTest)
library(buildmer)
library(geepack)
library(ggstatsplot)

df_meth <- readr::read_csv(here("raw_data", "methylation.csv"), col_types = "innnff")

df_eliktu <- readr::read_csv(here("raw_data", "df_lapsed.csv")) %>%
  dplyr::filter(!rowSums(is.na(.)) >= 64) %>% 
  dplyr::rowwise() %>% 
  dplyr::mutate(aImp = sum(c(AMIS3, (6-AMIS7), AMIS11, AMIS15, AMIS19, AMIS23,
                             AMIS2, (6-AMIS6), AMIS10, AMIS14, AMIS18, AMIS22)),
                mImp = sum(c(AMIS1, AMIS5, AMIS9, AMIS13, AMIS17, (6-AMIS21),
                             AMIS4, AMIS12, AMIS16, AMIS20, AMIS24))) %>%
  dplyr::ungroup() %>%
  dplyr::select(-c(AMIS1:AMIS24)) %>% 
  dplyr::filter(!(kood == 1824 & age == 18)) %>% 
  dplyr::filter(!(kood == 1157 & age == 33)) %>% 
  dplyr::filter(!(kood == 1307 & age == 18)) %>%
  dplyr::select(-c(sugu:kohort), -c(aImp:mImp), -LDL, -Wmaxkg) 
df_meth_tall <- df_meth |> tidyr::pivot_longer(cols = c(-kood, -MAOA.uVNTR, -Maltreatment.18),
                                               names_to = "age",
                                               values_to = "CpG3.methylation",
                                               names_prefix = "CpG3.methylation.",
                                               names_transform = as.integer)

df <- df_meth_tall |> dplyr::left_join(df_eliktu, by = c("kood", "age")) %>%
  dplyr::filter(!rowSums(is.na(.)) >= 34)
  

rowSums(is.na(df))

# saveRDS(df, here("data", "eliktu_lapsed_methylation_data.rds"), compress='xz')
# readr::write_csv2(df, here("data", "eliktu_lapsed_methylation_data.csv"))
```

```{r methyl-data}
skim(df_meth[, -c(1, 5)])

```

```{r correlations}

correlation(df_meth[, -c(1, 5)])
  
```

```{r ggpairs}
#| fig.height: 10

ggpairs(df_meth[, 2:4], mapping = aes(color = df_meth$MAOA.uVNTR))
```

```{r model-prep}
df_scaled  <- df %>% 
  dplyr::mutate(kood = factor(kood), age_cent = (age - 18), age = NULL) %>% 
  dplyr::mutate(across(CpG3.methylation:Alco, ~ c(scale(., scale = TRUE))))

# readr::write_csv2(df_scaled, here("data", "eliktu_lapsed_methylation_data_scaled.csv"))
```

```{r formula}

fmla <- reformulate(c(names(df_scaled)[-c(1, 3)], "(1 + age_cent | kood)"), names(df_scaled)[3])


```

**Finding the maximal *feasible* model & doing stepwise elimination from it**

The selected formula after the backward elimination `CpG3.methylation ~ 1 + Zink + age_cent`

The criterion for variable elimination is the significance of the change in log-likelihood (`LRT`: terms which provide lower chi-square values are considered more important)

`lmer` models do not converge even with only few variables, therefore `lm` was used.

```{r lmer}
m1 <- buildmer(fmla, data = df_scaled, 
         buildmerControl = buildmerControl(direction = 'backward',
          args = list(control = lmerControl(optimizer = 'bobyqa'))))

```

```{r model-summary}
summary(m1)

```

Using AIC as a variable selection criterion

```{r lmer2}
fmla2 <- reformulate(c(names(df_scaled)[-c(1, 3)], "(1 | kood)"), names(df_scaled)[3])
m2 <- buildmer(fmla2, data = df_scaled, 
         buildmerControl = buildmerControl(direction = c('order', 'backward'),
                                           crit = 'AIC',
          args = list(control = lmerControl(optimizer = 'bobyqa'))))
summary(m2)
```

glmmTMB

```{summary(m3){r glmmTMB}

m3 <- buildglmmTMB(fmla2, data = df_scaled, 
         buildmerControl = buildmerControl(direction = c('order', 'backward'),
                                           crit = 'AIC'))


```

There was no significant interaction between age and zinc.Then I added MAO-A genotype and maltreatment factors.

```{r age-zink}

fmla3 <- as.formula("CpG3.methylation ~ Zink + age_cent + Maltreatment.18 * MAOA.uVNTR")

m4 <- lm(fmla3, data = df_scaled)
summary(m4)
```

**Generalised estimating equations (GEE)**

GEE estimates population average effects.

```{r gee}

df_scaled2 <- df_scaled |> dplyr::mutate(age_cent = age_cent + 3L)

fmla4 <- formula(CpG3.methylation ~ BMI + Zink + Alco + Iron + age_cent +      Iodine + VitE + Fatsg + FruitsBerries + CHL + VitB1 + Protein +  
    Potassium + Phosphorus + Manganese + Calcium)
fmla4
m5 <- geeglm(fmla4, data = na.omit(df_scaled2), id = kood, family = gaussian, corstr= "ex", scale.fix = FALSE)
summary(m5)
```

MAO-A genotype and maltreatment at age 18

```{r maltr-18}

df_18 <- df |> dplyr::filter(age == 18L)

grouped_ggbetweenstats(
  data             = df_18,
  x                = Maltreatment.18,
  y                = CpG3.methylation,
  grouping.var     = MAOA.uVNTR,
  outlier.tagging  = TRUE,
  outlier.label    = kood,
  outlier.coef     = 2,
  ggsignif.args    = list(textsize = 4, tip_length = 0.01),
  p.adjust.method  = "bonferroni",
  palette          = "default_jama",
  package          = "ggsci",
  plotgrid.args    = list(nrow = 1),
  annotation.args  = list(title = "Differences in CpG3 methylation by maltreatment status at age 18 for MAO-A genotypes")
)

```

