------------------------------------------------------------------------

---
title: "Methylation"
format: html
editor: visual
knitr:
  opts_chunk: 
    cache : true
    cache.comments: false
    cache.lazy: false
    collapse: true
    warning: false
    message: false
    echo: false
    strip.white: true
    comment: "#>"
    fig.path: "../figures/"
    results: "asis"
    tidy: "styler"
    dev: "pdf"
    dev.args: list(pdf = list(colormodel = 'cmyk'), 
                  tiff = list(compression = 'lzw'))
    dpi: 600
    fig.width: 8
    fig.height: 8
    concordance: true
    global.par: true
    R.options:
      scipen: 999
      huxtable.long_minus: true
---

```{r load-data}

library(tidyverse)
library(here)
library(skimr)
library(correlation)
library(GGally)
library(lme4)
library(lmerTest)
library(buildmer)
library(ggstatsplot)

df_eliktu <- readr::read_csv(here("raw_data", "df_lapsed.csv")) %>%
  dplyr::filter(!rowSums(is.na(.)) >= 64) %>% 
  dplyr::rowwise() %>% 
  dplyr::mutate(aImp = sum(c(AMIS3, (6-AMIS7), AMIS11, AMIS15, AMIS19, AMIS23,
                             AMIS2, (6-AMIS6), AMIS10, AMIS14, AMIS18, AMIS22)),
                mImp = sum(c(AMIS1, AMIS5, AMIS9, AMIS13, AMIS17, (6-AMIS21),
                             AMIS4, AMIS12, AMIS16, AMIS20, AMIS24))) %>%
  dplyr::ungroup() %>%
  dplyr::select(-c(AMIS1:AMIS24)) %>% 
  dplyr::filter(!(kood == 1824 & age == 18)) %>% 
  dplyr::filter(!(kood == 1157 & age == 33)) %>% 
  dplyr::filter(!(kood == 1307 & age == 18)) %>%
  dplyr::select(-c(sugu:kohort), -c(aImp:mImp), -LDL, -Wmaxkg) %>%
  dplyr::mutate(kood = factor(kood))

df_police <- readr::read_csv(here("raw_data", "police_contact.csv"), col_types = "fii") %>% tibble::add_column(age = 33L) 

df_meth <- readr::read_csv(here("raw_data", "methylation.csv"), col_types = "innnff")

df_meth_tall <- df_meth |> tidyr::pivot_longer(cols = c(-kood, -MAOA.uVNTR, -Maltreatment.18),
                                               names_to = "age",
                                               values_to = "CpG3.methylation",
                                               names_prefix = "CpG3.methylation.",
                                               names_transform = as.integer) |> 
  dplyr::mutate(kood = factor(kood))




```

```{r scaled}
# recode age to join wth methylation dataset

df_33 <- df_police %>% dplyr::left_join(df_eliktu, by = c("kood", "age")) %>%
  dplyr::filter(!rowSums(is.na(.)) >= 34) %>% 
  dplyr::mutate(age = dplyr::recode(age, `33` = 25L)) %>%
  left_join(df_meth_tall, by = c("kood", "age")) %>%
  dplyr::filter(!rowSums(is.na(.)) >= 34)

df_33_scaled <- df_33 %>% dplyr::mutate(age = NULL) %>% 
  dplyr::mutate(dplyr::across(c(Calcium:Alco, CpG3.methylation), ~ c(scale(., scale = TRUE))))


```

```{r formula}

fmla <- reformulate(names(df_33_scaled)[-c(1:3)], names(df_33_scaled)[2])

```

m1 \<- buildmer(fmla, data = df_33_scaled,

buildmerControl = buildmerControl(direction = c('order', 'backward'),

crit = 'AIC', family = binomial))

summary(m1)

```{r glm}

m1 <- buildmer(fmla, data = df_33_scaled, 
         buildmerControl = buildmerControl(direction = c('order', 'backward'),
                                           crit = 'AIC', family = binomial))

summary(m1)

fmla2 <- formula("Police_contact ~ 1 + Manganese + 
    VitE + Maltreatment.18 + Iron + SugSweets + BMI +
    Cerealprod + Protein + Niacin + Potassium + VitB6 + 
    CpG3.methylation + MAOA.uVNTR")

m1a <- glm(fmla2, data = df_33_scaled, family = binomial)

summary(m1a)
```

df_police_25 \<- df_police \|\> dplyr::mutate(age = dplyr::recode(age, \`33\` = 25L))

df_25 \<- df_police_25 %\>%

dplyr::left_join(df_eliktu, by = c("kood", "age")) %\>%

left_join(df_meth_tall, by = c("kood", "age")) %\>%

dplyr::filter(!rowSums(is.na(.)) \>= 34)

df_25_scaled \<- df_25 %\>% dplyr::mutate(age = NULL) %\>%

dplyr::mutate(across(c(Calcium:Alco, CpG3.methylation), \~ c(scale(., scale = TRUE))))

```{r age-25}

df_police_25 <- df_police |> dplyr::mutate(age = dplyr::recode(age, `33` = 25L))

df_25 <- df_police_25 %>% 
  dplyr::left_join(df_eliktu, by = c("kood", "age")) %>%
  left_join(df_meth_tall, by = c("kood", "age")) %>%
  dplyr::filter(!rowSums(is.na(.)) >= 34)

df_25_scaled <- df_25 %>% dplyr::mutate(age = NULL) %>% 
  dplyr::mutate(across(c(Calcium:Alco, CpG3.methylation), ~ c(scale(., scale = TRUE))))

```

```{r glm-25}

# I use the same variables as in the model at age 33

m2 <- glm(fmla2, data = df_25_scaled, family = binomial)

summary(m2)
```

```{r age-18}

df_police_18 <- df_police |> dplyr::mutate(age = dplyr::recode(age, `33` = 18L))

df_18 <- df_police_18 %>% 
  dplyr::left_join(df_eliktu, by = c("kood", "age")) %>%
  left_join(df_meth_tall, by = c("kood", "age")) %>%
  dplyr::filter(!rowSums(is.na(.)) >= 34)

df_18_scaled <- df_18 %>% dplyr::mutate(age = NULL) %>% 
  dplyr::mutate(across(c(Calcium:Alco, CpG3.methylation), ~ c(scale(., scale = TRUE))))

```

```{r glm-18}
# I use the same variables as in the model at age 33

m3 <- glm(fmla2, data = df_18_scaled, family = binomial)

summary(m3)
```
